---
title: "Explore geographic variation"
subtitle: "suwo: access nature media repositories"
pagetitle: suwo map_locations vignette
author:
- <a href="https://marce10.github.io/">Marcelo Araya-Salas</a>, Jorge Elizondo & <a href="https://ecophysics.org/">Alejandro Rico-Guevara</a>
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    self_contained: false
    toc: true
    toc_depth: 2
    toc_float: true
editor_options: 
  chunk_output_type: console
bibliography: references.bib
link-citations: true
---

```{r setup, include=FALSE}
library(suwo)
knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  out.width = "100%",
  out.heigth = "100%"
)

```

```{r save queries, eval=FALSE, echo=FALSE}

# get metadata

### sound example ############
op_sound <- query_inaturalist(
  species = "Oophaga pumilio",
  format = "sound"
)

### image example ############
op_image <- query_inaturalist(
  species = "Oophaga pumilio",
  format = "image"
)

# restrict only to those in its natural range (Central America)
op_image_ca <- op_image[
  op_image$latitude >= 7 &
    op_image$latitude <= 15 &
    op_image$longitude >= -90 &
    op_image$longitude <= -80 &
    !is.na(op_image$latitude),
]


# sample homogeneously across the geographic range
# split latitude in 50 bins
op_image_ca$latitude_bins <- cut(
  op_image_ca$latitude,
  breaks = seq(7, 15, length.out = 50)
)

# sampled keys
sampled_keys <- op_image_ca$key[!duplicated(op_image_ca$latitude_bins)]
set.seed(123)
op_image_ca_sample <- op_image_ca[op_image_ca$key %in% sampled_keys, ]

### video example ############
cm_video <- query_macaulay(
  species = "Ceratopipra mentalis",
  format = "video",
  all_data = TRUE
)

cm_video_display <- cm_video[
  grepl("display", cm_video$behavior, ignore.case = TRUE),
]

## multiple species example ###########
alouatta_sound <- query_xenocanto("gen:Alouatta")

merged_data <- merge_metadata(
  op_sound,
  op_image_ca_sample,
  cm_video_display,
  alouatta_sound
)

# write.csv(merged_data, "./inst/articles/map_location_vignette_data.csv", 
#row.names = FALSE)

```


```{r load queries, eval=TRUE, echo=FALSE}

merged_data <- read.csv("map_location_vignette_data.csv")

op_sound <- merged_data[
  merged_data$format == "sound" & merged_data$species == "Oophaga pumilio",
]


op_image_ca_sample <- merged_data[merged_data$format == "image", ]

cm_video_display <- merged_data[merged_data$format == "video", ]

alouatta_sound <- merged_data[grep("Alouatta", merged_data$species), ]

```


::: {.alert .alert-info}

The goal of this vignette is to illustrate how to use the [map_locations()](https://docs.ropensci.org/suwo/reference/map_locations.html) function to explore the geographic distribution of media records as a proxy to understand geographic variation in biological traits.

:::

Understanding the geographic distribution of biological variation is essential not only for ecological and evolutionary research [@zamudio2016] but also for building robust automated classification models that account for phenotypic diversity [@pagano2023]. The [map_locations()](https://docs.ropensci.org/suwo/reference/map_locations.html) function in **suwo** allows researchers to interactively explore the spatial distribution of media records retrieved through the package, enabling rapid visual assessment of geographic patterns in diverse trait data.

The function accepts a metadata data frame —typically obtained from any of the `query_()` functions or as the output of [merge_metadata()](https://docs.ropensci.org/suwo/reference/merge_metadata.html) and [remove_duplicates()](https://docs.ropensci.org/suwo/reference/remove_duplicates.html) — and generates an interactive map powered by the [leaflet package](https://CRAN.R-project.org/package=leaflet). Each point on the map represents a unique observation with valid geographic coordinates. Clicking on a point opens a popup window displaying the associated media file(s) and selected metadata fields.

In this vignette, we demonstrate how to leverage the key features of [map_locations()](https://docs.ropensci.org/suwo/reference/map_locations.html) through three case studies: mapping image records of a color‑polymorphic frog, mapping sound recordings of the same species, and mapping video recordings of a bird's courtship display.

# Mapping images

The [Strawberry Poison Dart Frog (*Oophaga pumilio*)](https://en.wikipedia.org/wiki/Strawberry_poison_dart_frog) exhibits remarkable geographic variation in coloration across its range in Central America [@wang2008]. Populations display striking differences in dorsal and ventral coloration, ranging from bright red with blue legs to green, yellow, or even white morphs. This system provides an excellent opportunity to demonstrate how [map_locations()](https://docs.ropensci.org/suwo/reference/map_locations.html) can be used to explore spatial patterns of phenotypic variation using image records.

First, we query for image records of this species from iNaturalist:

```{r, eval = FALSE}
# get metadata
op_image <- query_inaturalist(
  species = "Oophaga pumilio",
  format = "image"
)
```

The raw query returns all available image records, including some that fall outside the species' natural range (e.g., observations from zoos, captive specimens, or misidentified individuals from other regions). To focus our analysis on wild populations within their native distribution, we restrict the dataset to records from Central America. Additionally, to keep the map manageable for this vignette and avoid overcrowding, we sample observations evenly across the geographic range by splitting the latitudinal range in 50 bins and selecting one record per bin. Note that because some observations have multiple associated images, we sample at the level of observation keys:

```{r, eval = FALSE}

# restrict only to those in its natural range (Central America)
op_image_ca <- op_image[
  op_image$latitude >= 7 &
    op_image$latitude <= 15 &
    op_image$longitude >= -90 &
    op_image$longitude <= -80 &
    !is.na(op_image$latitude),
]

# sample homogeneously across the geographic range
# split latitude in 50 bins
op_image_ca$latitude_bins <- cut(
  op_image_ca$latitude,
  breaks = seq(7, 15, length.out = 50)
)

# sampled keys
sampled_keys <- op_image_ca$key[!duplicated(op_image_ca$latitude_bins)]
set.seed(123)
op_image_ca_sample <- op_image_ca[op_image_ca$key %in% sampled_keys, ]

# make map
map_locations(
  metadata = op_image_ca_sample
)
```

```{r, echo = FALSE}

# make map
map_locations(
  metadata = op_image_ca_sample
)

```

When users click on any observation circle, a popup window appears displaying the image(s) associated with that record along with its most relevant metadata. This functionality allows researchers to visually inspect the color morph present at each locality, enabling rapid assessment of geographic patterns in coloration. For observations that contain multiple images (e.g., different angles or life stages), users can navigate between them using the arrow buttons at the bottom of the popup. By default the metadata includes the observation key, which contains a link to the original record on the repository, which can be used to explore additional information about the record. Users can download the image by simply clicking on them.


```{r show-southern-popup, echo=FALSE}
map_locations_preview <- function(metadata, ...) {

  metadata <- metadata[!is.na(metadata$latitude) & !is.na(metadata$longitude), ]

  # calculate centroid (mean of all coordinates)
  centroid_lat <- mean(metadata$latitude, na.rm = TRUE)
  centroid_lon <- mean(metadata$longitude, na.rm = TRUE)

  # calculate distance from each point to centroid
  distances <- sqrt((metadata$latitude - centroid_lat)^2 + 
                     (metadata$longitude - centroid_lon)^2)
  
  # find index of point closest to centroid
  idx <- which.min(distances)

  target_lat <- metadata$latitude[idx]
  target_lon <- metadata$longitude[idx]

  leaf_map <- map_locations(metadata, ...)

  leaf_map <- htmlwidgets::onRender(
    leaf_map,
    sprintf("
    function(el, x) {

      var map = this;

      map.whenReady(function() {

        setTimeout(function() {

          var targetLat = %f;
          var targetLon = %f;

          map.eachLayer(function(layer) {

            if (layer.getLatLng && layer.getPopup) {

              var latlng = layer.getLatLng();

              if (
                Math.abs(latlng.lat - targetLat) < 1e-6 &&
                Math.abs(latlng.lng - targetLon) < 1e-6
              ) {
                layer.openPopup();
              }

            }

          });

        }, 500);

      });

    }
    ",
    target_lat, target_lon)
  )

  leaf_map
}
```

```{r, echo=FALSE}
map_locations_preview(
  op_image_ca_sample,
  popup_size = 0.8
)
```

## Customizing popup appearance

The `popup_size` argument controls the scaling of the popup size. For example, setting `popup_size = 0.6` displays popups at 60% of its original size, which can be useful when working with very large images or when wanting to see more context around each point on the map:

```{r, eval=FALSE}
map_locations(
  metadata = op_image_ca_sample,
  popup_size = 0.6
)
```


```{r, echo=FALSE}
map_locations_preview(
  op_image_ca_sample,
  popup_size = 0.6
)
```

## Controlling displayed metadata

The `tags` argument allows users to control which metadata fields are displayed in the popup. By default, [map_locations()](https://docs.ropensci.org/suwo/reference/map_locations.html) shows a set of commonly useful fields (e.g., repository, observation key, species, date, country, locality, and user name). However, this behavior can be customized to include only the variables relevant to a given research question:

```{r, eval=FALSE}
map_locations(
  metadata = op_image_ca_sample,
  tags = c( "key", "date"),
  popup_size = 0.6
)
```

```{r, echo=FALSE}
map_locations_preview(
  op_image_ca_sample,
  tags = c( "key", "date"),
  popup_size = 0.6
)
```

... or even suppress all metadata by setting `tags = NULL`:
```{r, eval=FALSE}
map_locations(
  metadata = op_image_ca_sample,
  tags = NULL,
  popup_size = 0.6
)
```

```{r, echo=FALSE}
map_locations_preview(
  op_image_ca_sample,
  tags = NULL,
  popup_size = 0.6
)
```

The map also includes a toggle button (located in the top-right corner) that can open all popups simultaneously, which is particularly useful for quickly scanning multiple records:

```{r show-all-popups, echo=FALSE}
map_locations_popups_on <- function(metadata, ...) {
  leaf_map <- map_locations(metadata, ...)
  leaf_map <- htmlwidgets::onRender(
    leaf_map,
    "
    function(el, x) {
      var map = this;
      map.whenReady(function() {
        function tryActivate() {

          var toggle = el.querySelector('[id^=\"popup-toggle-\"]');
          var knob = el.querySelector('[id^=\"toggle-knob-\"]');
          var sw = el.querySelector('[id^=\"toggle-switch-\"]');

          if (!toggle || !knob || !sw) {
            setTimeout(tryActivate, 200);
            return;
          }

          // activate popups by clicking the toggle
          toggle.click();

          // after popups exist, adjust view
          setTimeout(function() {
            var bounds = [];
            map.eachLayer(function(layer) {
              if (layer.getLatLng && layer.getPopup && layer.getPopup()) {
                var ll = layer.getLatLng();
                bounds.push([ll.lat, ll.lng]);
              }
            });
            if (bounds.length > 0) {
              map.fitBounds(bounds, {padding: [30, 30]});
            }
          }, 300);
        }
        setTimeout(tryActivate, 400);
      });
    }
    "
  )
  leaf_map
}
```

```{r, echo=FALSE}
map_locations_popups_on(
  op_image_ca_sample,
  tags = NULL,
  popup_size = 0.3
)
```

# Mapping sound recordings

Beyond visual traits, vocalizations often show geographic variation that can reveal patterns of cultural evolution, cryptic diversity, and population structure. The [Strawberry Poison Dart Frog](https://en.wikipedia.org/wiki/Strawberry_poison_dart_frog) is a great example: its advertisement calls seem to be much less variables across its range, providing an interesting contrast with its striking geographic variation in coloration. With [map_locations()](https://docs.ropensci.org/suwo/reference/map_locations.html), researchers can map where recordings come from and directly compare calls across different localities:

```{r, eval = FALSE}

op_sound <- query_inaturalist(
  species = "Oophaga pumilio",
  format = "sound"
)

# make map
map_locations(
  op_sound,
  tags = NULL,
  popup_size = 0.6
)

```

```{r, echo=FALSE}
map_locations_preview(
  op_sound,
  popup_size = 0.7
)
```

When clicking on a point representing a sound recording, the popup displays an embedded audio player that allows users to listen to the recording directly within the map interface. This feature is particularly valuable for bioacoustic research, as it enables rapid comparison of vocalizations across geographic space without downloading files or switching between applications. The audio player supports standard playback controls including play, pause, and volume adjustment and allows users to directly download the audio file (although this might depend on the browser).

# Mapping video recordings

Video recordings capture dynamic behaviors that cannot be fully appreciated through static images or audio alone. Courtship displays, locomotion, feeding behavior, and other complex activities are often best documented through video. To illustrate this functionality, we examine video recordings of the [Red-capped Manakin (*Ceratopipra mentalis*)](https://en.wikipedia.org/wiki/Red-capped_manakin), a species known for its elaborate courtship displays involving rapid wing-snapping, moonwalking, and coordinated jumps:

```{r, eval = FALSE}

cm_video <- query_macaulay(
  species = "Ceratopipra mentalis",
  format = "video",
  all_data = TRUE
)

# Filter for records containing display behavior
cm_video_display <- cm_video[
  grepl("display", cm_video$behavior, ignore.case = TRUE),
]

# make map
map_locations(
  cm_video_display,
  tags = NULL,
  popup_size = 0.6
)
```

```{r, echo=FALSE}
map_locations_popups_on(
  cm_video_display,
  tags = NULL,
  popup_size = 0.7
)
```

Video popups include an embedded video player that supports playback directly within the map and allows users to directly download the audio file (although this might depend on the browser). Note that video files are typically larger than images or audio files, so loading times may vary depending on internet connection speed and file size.

# Advanced customization options

The [map_locations()](https://docs.ropensci.org/suwo/reference/map_locations.html) function offers several additional arguments for customizing map appearance and behavior.

## Location markers

By default, observations are displayed as semi-transparent circles, which work well for visualizing density and avoiding visual clutter. However, users can instead display observations as markers using the `type = "markers"` argument:

```{r}

# make map
map_locations(
  op_image_ca_sample,
  type = "markers",
  marker_color = "darkblue",
)
```


## Marker clustering

When working with large datasets or when multiple observations occur in close proximity, markers can overlap and become difficult to distinguish. The `cluster = TRUE` argument enables marker clustering, which groups nearby markers into a single cluster marker that displays the number of observations it contains. Clicking on a cluster zooms in to reveal the individual markers, providing a cleaner map interface while preserving access to all data points:

```{r}

# make map
map_locations(
  op_image_ca_sample,
  type = "markers",
  marker_color = "darkblue",
  cluster = TRUE
)
```

## Color customization

The `marker_color` argument allows users to change the color of markers when `type = "markers"`. For circle-based maps (`type = "circles"`), colors are determined by the `palette` argument, which accepts any color palette.

## Coloring by categorical variables

The `by` argument allows users to color observations according to a categorical variable in the metadata. In this example we search for sound recordings of [howler monkey species (genus *Alouatta*)](https://en.wikipedia.org/wiki/Howler_monkey) from Xenocanto and use `by = "species"` to color points according to species identity:

```{r, eval = FALSE}

# search for species of the genus manacus
alouatta_sound <- query_xenocanto("gen:Alouatta")

# make map
map_locations(
  alouatta_sound,
  by = "species",
  palette = grDevices::hcl.colors # viridis palette
)

```

```{r, eval = TRUE, echo = FALSE}

# make map
map_locations(
  alouatta_sound,
  by = "species",
  palette = grDevices::hcl.colors
)

```

## Suppressing media display

For situations where only metadata information is needed, users can suppress media display entirely using `show_media = FALSE`. This creates popups that contain only the metadata specified in the `tags` argument, without any audio players, images, or video players, resulting in faster rendering and reduced bandwidth usage.

```{r, eval = FALSE, echo = TRUE}

# make map
map_locations(
  alouatta_sound,
  by = "species", 
  show_media = FALSE,
  palette = grDevices::hcl.colors
)

```

```{r, eval = TRUE, echo = FALSE}

# make map
map_locations_preview(
  alouatta_sound,
  by = "species", 
  palette = grDevices::hcl.colors,
  )

```


## Further customization of leaflet maps

The function returns a `leaflet` map object, which can be further customized using any of the options available in the [leaflet](https://CRAN.R-project.org/package=leaflet) and [leaflet.extras](https://CRAN.R-project.org/package=leaflet.extras) packages. Below is an example of how to add satellite imagery and OpenStreetMap tiles as basemap options, allowing users to toggle between them:

```{r}

op_map <- map_locations(
  metadata = alouatta_sound,
  tags = c("key", "species"),
  popup_size = 0.6,
  palette = grDevices::hcl.colors # viridis palette  
)

op_map |>
  leaflet::addProviderTiles(
    leaflet::providers$Esri.WorldImagery,
    group = "Satellite"
  ) |>
  leaflet::addProviderTiles(leaflet::providers$OpenStreetMap, group = "Map") |>
  leaflet::addLayersControl(
    baseGroups = c("Satellite", "Map"),
    options = leaflet::layersControlOptions(collapsed = FALSE),
    position = "topleft"
  )
```


Check the [leaflet package documentation](https://rstudio.github.io/leaflet/index.html) and the [leaflet.extras package](https://CRAN.R-project.org/package=leaflet.extras) for more information on how to customize maps.

### References {.unnumbered .unlisted}

<div id="refs"></div>

<!-- add packages used, system details and versions  -->


### Session information {.unnumbered .unlisted}

<details>
  <summary>Click to see</summary>
```{r session info}
#| echo: false

sessionInfo()

```
</details>

