---
title: "Xeno-Canto annotations"
subtitle: "suwo: access nature media repositories"
pagetitle: suwo Xeno-Canto annotations vignette
author:
- <a href="https://marce10.github.io/">Marcelo Araya-Salas</a>
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    self_contained: false
    toc: true
    toc_depth: 2
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(suwo)
knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  out.width = "100%",
  out.heigth = "100%"
)

```

```{r save queries, eval=FALSE, echo=FALSE}

# Xeno-Canto's API can be unreliable (intermittent server errors), so
# queries for this vignette are run ONCE here and their results cached to
# an RDS file. Every subsequent render loads from that file instead of
# hitting the live API (see the "load queries" chunk below). To refresh
# the cache, run this chunk manually, then uncomment and run the
# saveRDS() line.
#
# NOTE: only the RESULT objects are cached, not the printed console
# messages. Those are instead shown below as plain, hand-written text
# matching real captured output. Live-replaying suwo's actual message()
# output turned out to be fragile across rendering contexts (raw
# terminal escape sequences, custom knitr hooks being bypassed, and
# angle-bracket text like "<your_object>" being silently swallowed when
# interpreted as HTML), so static text sidesteps all of that. If
# suwo's message wording changes, update the text blocks below by hand
# to match.

query_list <- list(
  pippip = query_xenocanto(species = "Pipistrellus pipistrellus", pb = FALSE),
  corcor = query_xenocanto(species = 'ann_sp:"Pelophylax ridibundus"', pb = FALSE),
  grycamp = query_xenocanto(
    species = 'grp:grasshoppers ann_sp:"Gryllus campestris"',
    pb = FALSE
  )
)

# just make sure this works

  all_bats <- query_xenocanto(species = "ann:yes grp:bats", pb = FALSE)
  all_frogs <- query_xenocanto(species = "ann:yes grp:frogs", pb = FALSE)

# saveRDS(query_list, "./vignettes/xenocanto_annotations_queries.rds")

```

```{r load queries, eval=TRUE, echo=FALSE}

query_list <- readRDS("xenocanto_annotations_queries.rds")

pippip <- query_list$pippip
corcor <- query_list$corcor
grycamp <- query_list$grycamp

```

::: {.alert .alert-info}

The goal of this vignette is to illustrate how to query for and retrieve Xeno-Canto annotation data using [query_xenocanto()](https://docs.ropensci.org/suwo/reference/query_xenocanto.html)

:::

Xeno-Canto recordings are sometimes annotated: one or more sounds in the sound file are individually marked with their own species identification, time and frequency bounds, sound type, and other details. 

suwo (version >= 0.2.3) now allows users to retrieved annotations along with the associated sound files. Whenever the recordings matching a search have annotations attached, they are automatically extracted and made available as an attribute of the result.

# Basic usage

[query_xenocanto()](https://docs.ropensci.org/suwo/reference/query_xenocanto.html) requires a Xeno-Canto API key. Get yours at [https://xeno-canto.org/account](https://xeno-canto.org/account), then set it as an environment variable named `xc_api_key`:

```{r, eval = FALSE}
# run this in the console, don't save it in a script
Sys.setenv(xc_api_key = "YOUR_API_KEY_HERE")
```

A regular call to [query_xenocanto()](https://docs.ropensci.org/suwo/reference/query_xenocanto.html) works exactly as usual:

```r
pippip <- query_xenocanto(species = "Pipistrellus pipistrellus")
```
```
ℹ Obtaining metadata: 
✔ 1154 matching sound files found 🎊
✔ 2 annotations found and added to the returned object's `annotations` attribute (access with attr(<your_object>, "annotations")) 🌈
```

The main result is the same data frame returned by any of the suwo's query function, one row per recording:

```{r}
head(pippip[, c("key", "species", "country", "locality")], 4)
```

Whenever any of the matching recordings have annotations attached, a second message reports how many were found and confirms they were added to the result as an attribute. If none of the matching recordings have annotations, this message is simply not shown, and the attribute is `NULL`.

# Accessing the annotations

Retrieve the annotations with `attr()`:

```{r}
annotations <- attr(pippip, "annotations")

nrow(annotations)
```

Each row is one annotated segment, not one recording. A single recording can have several annotations, each describing a different part of that one sound file:

```{r}
annotations[, c("key", "species", "start_time", "end_time", "frequency_low", "frequency_high")]
```

# Understanding the annotation columns

The full set of columns returned includes:

```{r}
names(annotations)
```

A few worth highlighting:

- **`key`**. The Xeno-Canto ID of the *recording* this annotation belongs to. This matches the `key` column of the main result, and is the column to join on (see below).
- **`species`**. The species identified in this specific annotated segment. This is usually, but not necessarily always, the same as the parent recording's overall species. It can differ for an annotated background call of a different species.
- **`start_time`** / **`end_time`**. The annotated segment's time bounds within the recording, in seconds.
- **`frequency_low`** / **`frequency_high`**. The annotated segment's frequency bounds, in Hz.
- **`annotator`**. Who created this specific annotation.
- **`sound_type`**, **`sex`**, **`life_stage`**, **`annotation_remarks`**. Further details about the annotated segment, when available.
- **`annotation_set_name`**, **`annotation_set_creator`**, **`annotation_set_creation_date`**, **`annotation_set_remarks`**. Metadata about the *set* of annotations this one belongs to (a recording's annotations are usually created together, as one set).
- **`file_url`** / **`observation_url`**. Links back to the parent recording.

Since several annotation rows commonly share the same `key` (several segments from the same recording), values in `key` and `species` are expected to repeat. This reflects the data correctly rather than indicating a problem.


# Searching specifically for annotated recordings

Annotation data is available on any query's result whenever present, regardless of the search used. To instead search specifically for recordings that have an annotation matching certain criteria, use Xeno-Canto's `ann_*` search tags directly in the `species` argument, e.g. `ann_sp` for the annotated species, `ann_type` for the annotated sound type:

```r
# recordings with an annotation identified as this species
corcor <- query_xenocanto(species = 'ann_sp:"Corvus corax"')
```
```
ℹ Obtaining metadata: 
✔ 153 matching sound files found 🎊
✔ 491 annotations found and added to the returned object's `annotations` attribute (access with attr(<your_object>, "annotations")) 😸
```

This only changes *which recordings* are returned by the search,  but does not change how annotation data itself is retrieved or structured. The `annotations` attribute is extracted the same way either way.

## Landscape recordings

Some annotations come from **soundscape/landscape recordings** rather than a recording of a single focal species. Xeno-Canto labels the annotated species in these cases as `Sonus naturalis` (not a real biological taxon, but a placeholder Xeno-Canto uses specifically for this recording type). This can show up unexpectedly when inspecting annotation species, even for a search targeting a real species:

```{r}
table(attr(corcor, "annotations")$species)
```

Keep this in mind when filtering or summarizing annotations by species. `Sonus naturalis` entries describe general soundscape content picked up in the background, not an actual species identification for that segment. To search specifically for this kind of recording, use `grp:soundscape` (or `grp:0`). See [Searching non-bird taxa](#searching-non-bird-taxa) below.

## Searching non-bird taxa

The `grp:` tag narrows a search to a specific taxonomic group, and is especially useful combined with other tags. Valid values are `grp:birds`, `grp:grasshoppers`, `grp:bats`, `grp:frogs`, and `grp:"land mammals"` (note the quotes, needed because the value contains a space). Numeric IDs work too (`1` through `5` for the five groups above, e.g. `grp:2` is equivalent to `grp:grasshoppers`). Soundscape recordings are a special case, since they can include multiple groups at once. Use `grp:soundscape` or `grp:0` to search those specifically.

Including `grp:` matters in particular when combined with `ann_*` annotation tags: omitting it for a non-bird species alongside an annotation tag can cause the request to fail outright with a server error, rather than simply returning no results:

```{r, eval = FALSE}
# fails with an HTTP 500 server error (grasshoppers require grp: to be
# specified explicitly when searching by annotation tags)
query_xenocanto(species = 'ann_sp:"Gryllus campestris"')
```

Including `grp:` resolves it:

```r
grycamp <- query_xenocanto(
  species = 'grp:grasshoppers ann_sp:"Gryllus campestris"'
)
```
```
ℹ Obtaining metadata: 
✔ 4 matching sound files found 🎊
✔ 10 annotations found and added to the returned object's `annotations` attribute (access with attr(<your_object>, "annotations")) 🌈
```

The `ann:yes` tag searches for recordings that have *any* annotation attached, regardless of what species or type it identifies. This is unlike the `ann_*` tags (`ann_sp`, `ann_type`, etc.), which filter by a specific annotation attribute. Combined with `grp:`, this is a convenient way to browse annotated recordings across a whole non-bird group:

```r
all_bats <- query_xenocanto(species = "ann:yes grp:bats")

all_frogs <- query_xenocanto(species = "ann:yes grp:frogs")
```

Annotations can then be easily formatted to be used by other R packages like [warbleR](https://CRAN.R-project.org/package=warbleR) or [Rraven](https://CRAN.R-project.org/package=Rraven) for further manipulation and analysis.

For more details on Xeno-Canto's annotation system, see their [official documentation](https://xeno-canto.org/article/318).


### Session information {.unnumbered .unlisted}

<details>
  <summary>Click to see</summary>
```{r session info}
#| echo: false

sessionInfo()

```
</details>
